Inferring Galactic Parameters from Chemical Abundances with Simulation-Based Inference

### 2025-02-17 src/tex/ms.tex
[showyourwork] overleaf sync

In order to establish our new method based on SBI we need two ingredients: A simulator (in our case a GCE model) that simulates observational data from a set of model parameters (in our case the IMF slope and the Type Ia supernovae normalization) and a flexible way of parametrizing the posterior density conditioned on the observation in order to perform our inference. In the next subsections we describe both ingredients in detail.

@@

$\log_{10}\left(N_\mathrm{Ia}\right)$ & Number of SN\,Ia per $\mathrm{M}_\odot$ over 15\,Gyr & $-2.75\pm0.3$ & \citep[Tab.1\,]{2012PASA...29..447M}\\

@@

\paragraph{Nucleosynthetic yield tables}
We adopt the same nucleosynthetic yield tables as in \citep{Philcox_2019}, see their Sec.~2.2 for more details.
To test our method, we aim further at inferring parameters from a sample of stars taken from a hydrodynamical simulation of a MW type galaxy which we take from the IllustrisTNG project \citep{Pillepich2018}. To ensure maximal compatibility with TNG, we adopt their nucleosynthetic yield tables in \texttt{CHEMPY}, for enrichment by SN\,Ia, SN\,II and AGB stars. The utilized yields are summarized in Tab.\,\ref{tab:chempy_TNG_yields}, matching \citet[Tab.\,2]{2018MNRAS.473.4077P}, and we note that the SN\,II yields are renormalized such that the IMF-weighted yield ratios at each metallicity are equal to those from the \citet{2006ApJ...653.1145K} mass range models alone. \texttt{CHEMPY} uses only net yields, such that they provide only newly synthesized material, with the remainder coming from the initial SSP composition. These tables may not well-represent true stellar chemistry, and the effects of this mismatch are examined in Sec.\,\ref{subsec:mocks_wrong_yield} by performing inference using an alternative set of yields that does not match the yield set of the training data. For the analysis of observational data, we would want to use the most up-to-date yields, such as \citet{2016ApJ...825...26K} AGB yields, and carefully choose elements which are known to be well reproduced by our current models (e.g. shown by \citet{2019ApJ...874..102W,2019arXiv190806113G}), though this is not appropriate in our context. To facilitate best comparison with TNG, we further set the maximum SN\,II mass as $100\,\mathrm{M}_\odot$ (matching the IMF upper mass limit), adopt stellar lifetimes from \citet{portinari} and do not allow for any `hypernovae' \citep[in contrary to][]{2018ApJ...861...40P}).

@@

\begin{figure}[]
     \centering
     \includegraphics[width=\columnwidth]{figures/ape_NN.pdf}
     \vspace{-.5cm}
     \caption{Absolute percentage error of the NN emulator for the \texttt{CHEMPY} simulator. The orange histogram shows the full distribution of percentage errors with the vertical dashed line indicating the median and the vertical dotted lines indicating the first and third quartile. The box plot on the top of the plot extends from the first quartile to the third quartile of the data, with a line at the median. The whiskers extend from the box to the farthest data point lying within $1.5\times$ the inter-quartile range from the box. The NN predicts abundances with an absolute percentage error far below typical observational errors.}
     \label{fig:ape_NN}
     \script{train_torch_chempy.py}
\end{figure}

\paragraph{Bayesian model}
\texttt{CHEMPY} effectively is a Bayesian model for stellar abundances given a set of parameters $\{\Vec{\Lambda},\Vec{\Theta},T\}$ and \citet{Philcox_2019} extended the \texttt{CHEMPY} framework to be able to model multiple stellar data-points. Consider a given star with index $i$ that is born in some region of the ISM. This star will carry its own set of parameters $\{\Vec{\Lambda},\Vec{\Theta}_i,T_i\}$, where $\Vec{\Lambda}$ are star independent and hence taken to be global parameters while the ISM parameters $\Vec{\Theta}_i$ and the birth-time $T_i$ are star specific. Using \texttt{CHEMPY} (or the trained neural network emulator) we can easily model the set of $n_\mathrm{el}$ chemical abundances $\{X_i^j\}$ for the $i$-th star as: 
\begin{equation}\label{eq:chempy_function}
\{X_i^j\} = \texttt{CHEMPY}(\Vec{\Lambda},\Vec{\Theta}_i,T_i),
\end{equation}
where $j$ indexes the chemical element. 
These model abundances can then be compared against observations, with measured abundances $d_i^j$ and corresponding Gaussian errors $\sigma_{i,\mathrm{obs}}^j$ jointly denoted as $D_i=\{d_i^j,\sigma_{i,\mathrm{obs}}^j\}$.

As stated in \citet{Philcox_2019} the full posterior for this case is given by 
\begin{eqnarray}\label{eq:posterior}
    \mathbb{P}(\vec\Lambda,\{\vec\Theta_i\},\{T_i\}|\{D_i\}) &\propto&  \left[\prod_{i=1}^{n_\mathrm{star}}p_{\vec\Theta}(\vec\Theta_i)p_{T_i}(T_i)\right]
    \times p_{\vec\Lambda}(\vec\Lambda)\\
    \nonumber
    &\times& \mathcal{L}(\{D_i\}|\vec\Lambda,\{\vec\Theta_i\},\{T_i\})
\end{eqnarray}
where $p_V(V_i)$ is the prior on variable $V_i$ (belonging to the set $V$).

@@

We use a Neural Posterior Estimator (NPE) \cite{zeghal2022neuralposteriorestimationdifferentiable} which utilises the gradients of the generative model $\mathcal{M}$ with a Neural Spline Flow (NSF) \citet{durkan2019neuralsplineflows} containing 20 hidden features and 10 transformation layers for the normalizing flow. The NSF uses as invertible transformation smooth and flexible spline functions to allow efficient sampling. The expressivity of the spline allows the NDE to be capture complex distributions, while also maintaining computational tractability. The final model was selected after extensive hyperparameter tuning, varying: the architecture between NSF, Masked Autoregressive Flow (MAF) \cite{papamakarios2018maskedautoregressiveflowdensity}, and Mask Autoregressive 
Flow with rational-quadratic spline (MAF-RQS), the number of neurons between {10, 20, 50, 100}, and the number of transformations between {1, 5, 10, 30}, optimizing over the test set for both the highest mean log posterior probability and the best calibration as measured by the TARP value.

We train our NPE with $10^5$ data points consisting of $n_{\rm elements}=8$ simulated with the NN emulator described in the previous section with inputs sampled from the priors shown in Tab. \ref{tab:priors}. Training takes $\lesssim10$ minutes on an Intel(R) Xeon(R) Platinum 8468V with 192 CPU cores. 
We evaluate the accuracy of the NPE using $5\times10^4$ validation data points from the original \texttt{CHEMPY} simulator. In order to mimic observational noise, we add 5\% observational uncertainties to the abundances simulated with \texttt{CHEMPY} before feeding them to our NPE.
We have made sure that our NPE is well calibrated and posterior distributions are trustable. For more details on simulation-based calibration see appendix \ref{sec:sbc}.

@@

\paragraph{Multi-star inference}
Following equation~\ref{eq:posterior} we can calculate the joint posterior for a combined inference using abundance observations of multiple stars. For this, we will condition our NPE individually on single star observations and sample the posterior. Then, according to equation~\ref{eq:posterior} we simply have to multiply individual posteriors and account for the inverse weighting by some power of the prior which we take from table~\ref{tab:priors}.

This does however present one issue, that since we only have access to samples from the posterior and not the posterior itself it is difficult to evaluate equation~\ref{eq:posterior}. We circumvent this by approximating each single star posterior by a Gaussian and fit for the parameters of mean and covariance. With this it is straight forward to evaluate equation~\ref{eq:posterior} analytically. In fact, under our assumption the combined posterior is a product between the prior and the product of multiple Gaussians for the individual star posteriors. The latter product is also a Gaussian with mean $\mathbf{\mu'}$ and variance $\mathbf{\sigma'}$:
\begin{align}
    \mathbf{\mu'} &= \frac{\sum_{i=1}^{N_{stars}} \frac{\mu_i}{\sigma_i^2}}{\sum_{i=1}^{N_{stars}} \frac1{\sigma_i^2}} \\
\mathbf{\sigma'}^2 &= \frac1 {\sum_{i=1}^{N_{stars}} \frac1{\sigma_i^2}}
\end{align}

@@

\subsection{\texttt{CHEMPY} Mock Observational Data}
Our analysis uses mock observations drawn from the neural network emulator with fixed values of the global galactic parameters $\alpha_{\rm IMF}=-2.3$ and $\log_{10}(N_{\rm Ia})=-2.89$ and varying local parameters $\vec{\Theta}_i$ using priors from Tab. \ref{tab:priors}. Additionally we draw $T_i$ uniformly in the range $[2,12.8]$ Gyr to minimize overlap with the neural network training birth-time limits when observational uncertainties are included. 
Each set of parameters is passed to our \texttt{CHEMPY} emulator, producing eight true chemical abundances. In order to fully represent observational data, we augment this data with observational uncertainties by adding a Gaussian error of $0.05$ dex for the abundances representative for typical APOGEE data \citep{Majewski2016}. All our datasets and accompanying code to reproduce our data generation process and analysis is publicly available\footnote{\url{https://github.com/TobiBu/sbi-chempy}}.

@@

For each individual observation consisting of the abundance measurements of a single star we sample the posterior estimate for all six parameters $\{\vec\Lambda,\vec\Theta_i,T_i\}$ with 1,000 points. This takes around $0.3s$ for each star, making it extremely fast to infer the parameters of a large amount of stars. Our method takes around $10$ seconds to build a posterior for all six parameters for a dataset size of 1,000 stars (each time sampling the single star posterior with 1,000 points and fitting for the Gaussian parameters). Our combined runtime for the NPE training plus sampling is then 610 seconds for 1,000 stars hence our methods making it more than $1,180$ times faster than current HMC methods \citep[cf.][who need $40$h for only 200 stars]{Philcox_2019}. But since our approach is amortized and we do not need to retrain our NPE model each time we want to make an inference, we in fact achieve an inference speed-up of a factor of $\sim72,000$. \textcolor{red}{check if those speed-up values are correctly calculated...}
Hence, shorter computing times make it feasible to use orders of magnitudes more observations.

>>>>

In order to establish our new method based on SBI we need two ingredients: A simulator (in our case a GCE model) that simulates observational data from a set of model parameters (in our case the IMF slope and the Type Ia supernovae normalization) and a flexible way of parametrizing the posterior density conditioned on the observation in order to perform our inference (see Fig. \ref{fig:flowchart} for a schematic visual representation of our method). In the next subsections we describe both ingredients in detail.

@@

$\log_{10}\left(N_\mathrm{Ia}\right)$ & Number of SN\,Ia per $\mathrm{M}_\odot$ over 15\,Gyr & $-2.89\pm0.3$ & \citep[Tab.1\,]{2012PASA...29..447M}\\

@@

\paragraph{Nucleosynthetic yield tables}
We adopt the same nucleosynthetic yield tables as in \citep{Philcox_2019}, see their Sec.~2.2 for more details.
To test our method, we aim further at inferring parameters from a sample of stars taken from a hydrodynamical simulation of a MW type galaxy which we take from the IllustrisTNG project \citep{Pillepich2018}. To ensure maximal compatibility with TNG, we adopt their nucleosynthetic yield tables in \texttt{CHEMPY}, for enrichment by SN\,Ia, SN\,II and AGB stars. The utilized yields are summarized in Tab.\,\ref{tab:chempy_TNG_yields}, matching \citet[Tab.\,2]{2018MNRAS.473.4077P}, and we note that the SN\,II yields are renormalized such that the IMF-weighted yield ratios at each metallicity are equal to those from the \citet{2006ApJ...653.1145K} mass range models alone. \texttt{CHEMPY} uses only net yields, such that they provide only newly synthesized material, with the remainder coming from the initial SSP composition. These tables may not well-represent true stellar chemistry, and the effects of this mismatch are examined in Sec.\,\ref{subsec:mocks_wrong_yield} by performing inference using an alternative set of yields that does not match the yield set of the training data. For the analysis of observational data, we would want to use the most up-to-date yields, such as \citet{2016ApJ...825...26K} AGB yields, and carefully choose elements which are known to be well reproduced by our current models (e.g. shown by \citet{2019ApJ...874..102W,2019arXiv190806113G}), though this is not appropriate in our context. To facilitate best comparison with Ilustris TNG, we further set the maximum SN\,II mass as $100\,\mathrm{M}_\odot$ (matching the IMF upper mass limit), adopt stellar lifetimes from \citet{portinari} and do not allow for any `hypernovae' \citep[in contrary to][]{2018ApJ...861...40P}).

@@

\begin{figure}[]
     \centering
     \includegraphics[width=\columnwidth]{figures/ape_NN.pdf}
     \vspace{-.5cm}
     \caption{Cumulative absolute percentage error of the NN emulator for the \texttt{CHEMPY} simulator. The orange histogram shows the cumulative distribution of percentage errors with the vertical dashed line indicating the median and the vertical dotted lines indicating the first and third quartile. The box plot on the top of the plot extends from the first quartile to the third quartile of the data, with a line at the median. The whiskers extend from the box to the farthest data point lying within $1.5\times$ the inter-quartile range from the box. The NN predicts abundances with an absolute percentage error far below typical observational errors.}
     \label{fig:ape_NN}
     \script{train_torch_chempy.py}
\end{figure}

\subsection{Bayesian model}
\texttt{CHEMPY} effectively is a Bayesian model for stellar abundances given a set of parameters $\{\Vec{\Lambda},\Vec{\Theta},T\}$ and \citet{Philcox_2019} extended the \texttt{CHEMPY} framework to be able to model multiple stellar data-points. Consider a given star with index $i$ that is born in some region of the ISM. This star will carry its own set of parameters $\{\Vec{\Lambda},\Vec{\Theta}_i,T_i\}$, where $\Vec{\Lambda}$ are star independent and hence taken to be global parameters while the ISM parameters $\Vec{\Theta}_i$ and the birth-time $T_i$ are star specific. Using \texttt{CHEMPY} (or the trained neural network emulator) we can easily model the set of $n_\mathrm{el}$ chemical abundances $\{X_i^j\}$ for the $i$-th star as: 
\begin{equation}\label{eq:chempy_function}
\{X_i^j\} = \texttt{CHEMPY}(\Vec{\Lambda},\Vec{\Theta}_i,T_i),
\end{equation}
where $j$ indexes the chemical element. 
These model abundances can then be compared against observations, with measured abundances $d_i^j$ and corresponding Gaussian errors $\sigma_{i,\mathrm{obs}}^j$ jointly denoted as $D_i=\{d_i^j,\sigma_{i,\mathrm{obs}}^j\}$.

\paragraph{Posteriors for Galactic parameters}
As stated in \citet{Philcox_2019} the full posterior for this case is given by 
\begin{eqnarray}\label{eq:posterior}
    \mathbb{P}(\vec\Lambda,\{\vec\Theta_i\},\{T_i\}|\{D_i\}) &\propto&  \left[\prod_{i=1}^{n_\mathrm{star}}p_{\vec\Theta}(\vec\Theta_i)p_{T_i}(T_i)\right]
    \times p_{\vec\Lambda}(\vec\Lambda)\\
    \nonumber
    &\times& \mathcal{L}(\{D_i\}|\vec\Lambda,\{\vec\Theta_i\},\{T_i\})
\end{eqnarray}
where $p_V(V_i)$ is the prior on variable $V_i$ (belonging to the set $V$).

@@

We use a NPE \cite{zeghal2022neuralposteriorestimationdifferentiable} which utilises the gradients of the generative model $\mathcal{M}$ with a Neural Spline Flow (NSF) \citet{durkan2019neuralsplineflows} containing 20 hidden features and 10 transformation layers for the normalizing flow. The NSF uses as invertible transformation smooth and flexible spline functions to allow efficient sampling. The expressivity of the spline allows the NDE to be capture complex distributions, while also maintaining computational tractability. The final model was selected after extensive hyperparameter tuning, varying: the architecture between NSF, Masked Autoregressive Flow (MAF) \cite{papamakarios2018maskedautoregressiveflowdensity}, and Mask Autoregressive 
Flow with rational-quadratic spline (MAF-RQS), the number of neurons between {10, 20, 50, 100}, and the number of transformations between {1, 5, 10, 30}, optimizing over the test set for both the highest mean log posterior probability and the best calibration as measured by the TARP value \citep{Lemos2023}. For more details on TARP see \ref{sec:sbc} in the appendix.

We train our NPE with $10^5$ data points consisting of $n_{\rm elements}=8$ simulated with the NN emulator described in the previous section with inputs sampled from a uniform prior over the original $5\sigma$-range of the Gaussian priors shown in Tab. \ref{tab:priors} to provide better coverage of the parameter space. Training takes $\sim70$ minutes on an Intel(R) Xeon(R) Platinum 8468V with 192 CPU cores. 
We evaluate the accuracy of the NPE using $5\times10^4$ validation data points from the original \texttt{CHEMPY} simulator. In order to mimic observational noise, we add 5\% observational uncertainties to the abundances simulated with \texttt{CHEMPY} before feeding them to our NPE.
We have made sure that our NPE is well calibrated and posterior distributions are trustable. For more details on simulation-based calibration see appendix \ref{sec:sbc}.

@@

\paragraph{Multi-star inference}
Following eq.~\ref{eq:posterior} we can calculate the joint posterior for a combined inference using abundance observations of multiple stars. For this, we will condition our NPE individually on single star observations and sample the posterior. Then, according to eq.~\ref{eq:posterior} we simply have to multiply individual posteriors and account for the inverse weighting by some power of the prior which we take from table~\ref{tab:priors}.

This does however present one issue, that since we only have access to samples from the posterior and not the posterior itself it is difficult to evaluate eq.~\ref{eq:posterior}. We circumvent this by approximating each single star posterior by a Gaussian and fit for the parameters of mean and covariance. With this it is straight forward to evaluate eq.~\ref{eq:posterior} analytically. In fact, under our assumption the combined posterior is a product between the prior and the product of multiple Gaussians for the individual star posteriors. The latter product is also a Gaussian with mean $\mathbf{\mu'}$ and variance $\mathbf{\sigma'}$:
\begin{align}
    \mathbf{\mu'} &= \frac{\sum_{i=1}^{N_{stars}} \frac{\mu_i}{\sigma_i^2}}{\sum_{i=1}^{N_{stars}} \frac1{\sigma_i^2}} \\
\mathbf{\sigma'}^2 &= \frac1 {\sum_{i=1}^{N_{stars}} \frac1{\sigma_i^2}}
\end{align}

@@

\subsection{\texttt{CHEMPY} Mock Observational Data}
Our analysis uses mock observations drawn from the neural network emulator with fixed values of the global galactic parameters $\alpha_{\rm IMF}=-2.3$ and $\log_{10}(N_{\rm Ia})=-2.89$ and varying local parameters $\vec{\Theta}_i$ using priors from Tab. \ref{tab:priors}. Additionally we draw $T_i$ uniformly in the range $[2,12.8]$ Gyr to minimize overlap with the neural network training birth-time limits when observational uncertainties are included. 
Each set of parameters is passed to our \texttt{CHEMPY} emulator, producing eight true chemical abundances. In order to fully represent observational data, we augment this data with observational uncertainties by adding a Gaussian error of $0.05$ dex for the abundances representative for typical APOGEE data \citep{Majewski2016}.

@@

For each individual observation consisting of the abundance measurements of a single star we sample the posterior estimate for all six parameters $\{\vec\Lambda,\vec\Theta_i,T_i\}$ with 1,000 points. This takes around $0.3s$ for each star, making it extremely fast to infer the parameters of a large amount of stars. Our method takes around $50$ seconds to build a posterior for all six parameters for a dataset size of 1,000 stars (each time sampling the single star posterior with 1,000 points and fitting for the Gaussian parameters). Our combined runtime for the NPE training plus sampling is then 4250 seconds for 1,000 stars hence our methods making it more than $170$ times faster than current HMC methods \citep[cf.][who need $40$h for only 200 stars]{Philcox_2019}. But since our approach is amortized and we do not need to retrain our NPE model each time we want to make an inference, we in fact achieve an inference speed-up of a factor of $\sim14,400$.
Hence, shorter computing times make it feasible to use orders of magnitudes more observations.
